Fisheries example integrating FLR

### 2018-05-10 notebook/ms/SI4.Rmd
Minor edits, new read

Here we demonstrate how simulations in GMSE can be more fine-tuned to specific empirical situations through the use of `gmse_apply`. To do this, we use the same scenario described in [Supporting Information 3](SI3.pdf); we first recreate the basic scenario run in `gmse` using `gmse_apply`, and then build in additional modelling details including (1) [custom placement of user land](#land), (2) [parameterisation of individual user budgets](#budgets), and (3) [density-dependent movement of resources](#movement). We emphasise that these simulations are provided only to demonstrate the use of GMSE, and specifically to show the flexibility of the `gmse_apply` function, not to accurately recreate the dynamics of a specific system or make management recommendations.

@@

Additionally, we continue to use the following values for consistency, except in the case of `stakeholders`, where we reduce the number of farmers to `stakeholders = 8`. This is done to for two reasons. First, it speeds up simulations for the purpose of demonstration; second, it makes the presentation of landscape ownership easier (see below).

@@

We now recreate the simulations in [Supporting Information 3](SI3.pdf), which were run using the `gmse` function, in `gmse_apply`. Doing so requires us to first initialise simulations using one call of `gmse_apply`, then loop through multiple time steps that again call `gmse_apply`; results of interest are recorded in a data frame (`sim_sum_1`). Following the protocol introduced in [Supporting Information 2](SI2.pdf), we can call the initialising simulation `sim_old`, and use the code below to read in the relevant parameter values.

@@

We can then loop over 30 time steps to recreate the simulations from [Supporting Information 3](SI3.pdf). In these simulations, we are specifically interested in the resource and observation outputs, as well as the manager policy and user actions for culling, which we record below in the data frame `sim_sum`. The inclusion of the argument `old_list` tells `gmse_apply` to use parameters and values from the list `sim_old` in the new time step.

@@

The above output from `sim_sum_1` shows the data frame that holds the information we were interested in pulling out of our simulation results. All of this information was available under the list element `sim_new$basic_output`, but other list elements of `sim_new` might also be useful to record. It is important to remember that this example of `gmse_apply` is using the default resource, observation, manager, and user sub-models. Custom sub-models could produce different outputs in `sim_new` (see [Supporting Information 2](SI2.pdf) for examples). For default sub-models, there are some list elements that might be especially useful. These elements can potentially be edited *within the above loop* to dynamically adjust simulations.

- `sim_new$resource_array`: A table holding all information on resources. Rows correspond to discrete resources, and columns correspond to resource properties: (1) ID, (2-4) types (not currently in use), (5) x-location, (6) y-location, (7) movement parameter, (8) time, (9) density independent mortality parameter (`remove_pr`), (10) reproduction parameter (`lambda`), (11) offspring number, (12) age, (13-14) observation columns, (15) consumption rate (`res_consume`), and (16-20) recorded experiences of user actions (e.g., was the resource culled or scared?).
- `sim_new$AGENTS`: A table holding basic information on agents (manager and users). Rows correspond to a unique agent, and columns correspond to agent properties: (1) ID, (2) type (0 for the manager, 1 for users), (3-4) additional type options not currently in use, (5-6), x and y locations (usually ignored), (7) movement parameter (usually ignored), (8) time, (9) agent's viewing ability in cells (`agent_view`), (10) error parameter, (11-12) values for holding marks and tallies of resources, (13-15) values for holding observations, (16) yield from landscape cells, (17) budget (`manager_budget` and `user_budget`).
- `sim_new$observation_vector`: Estimate of total resource number from the observation model (`observation_array` also holds this information in a different way depending on `observe_type`)
- `sim_new$LAND`: The landscape on which interactions occur, which is stored as a 3D array with `land_dim_1` rows, `land_dim_2` columns, and 3 layers. Layer 1 (`sim_new$LAND[,,1]`) is not currently used in default sub-models, but could be used to store values that affect resources and agents. Layer 2 (`sim_new$LAND[,,2]`) stores crop yield from a cell, and layer 3 (`sim_new$LAND[,,3]`) stores the owner of the cell (value corresponds to the agent's ID).
- `sim_new$manage_vector`: The cost of each action as set by the manager. For even more fine-tuning, individual costs for the actions of each agent can be set for each user in `sim_new$manager_array`.
- `sim_new$user_vector`: The total number of actions performed by each user. A more detailed breakdown of actions by individual users is held in `sim_new$user_array`.

@@

The number of cells owned by each farmer is therefore listed in the table below.

@@

```{r, echo = TRUE, eval = TRUE}
# First initialise a simulation
sim_old  <- gmse_apply(get_res = "Full", remove_pr = 0.122, lambda = 0.275,
                       res_death_K = 93870, RESOURCE_ini = 35000, 
                       manage_target = 70000, res_death_type = 3,
                       manager_budget = 10000, user_budget = 10000,
                       public_land = 0.4, stakeholders = 8, res_consume = 0.02, 
                       res_birth_K = 200000, land_ownership = TRUE,
                       observe_type = 3, agent_view = 1, converge_crit = 0.01,
                       ga_mingen = 200, res_move_type = 0);
# By setting `res_move_type = 0`, no resource movement will occur in gmse_apply
# Adjust the landscape ownership below
sim_old$LAND[1:20,   1:20,  3] <- 2; 
sim_old$LAND[1:20,  21:40,  3] <- 3;
sim_old$LAND[1:20,  41:60,  3] <- 4;
sim_old$LAND[1:20,  61:80,  3] <- 5;
sim_old$LAND[1:20,  81:100, 3] <- 6;
sim_old$LAND[21:40,  1:50,  3] <- 7;
sim_old$LAND[21:40, 51:100, 3] <- 8;
sim_old$LAND[41:60,  1:100, 3] <- 9;
sim_old$LAND[61:100, 1:100, 3] <- 1;
# Change the budgets of each farmer based on the land they own
for(ID in 2:9){
    cells_owned            <- sum(sim_old$LAND[,,3] == ID);
    sim_old$AGENTS[ID, 17] <- 10 * cells_owned;
}
# Begin simulating time steps for the system
sim_sum_2 <- matrix(data = NA, nrow = 30, ncol = 5);
for(time_step in 1:30){
    # Apply the new movement rules at the beginning of the loop
    sim_old$resource_array  <- avoid_aggregation(sim_old$resource_array);
    # Next, move on to simulate (old_list remembers that res_move_type = 0)
    sim_new                 <- gmse_apply(get_res = "Full", old_list = sim_old);
    sim_sum_2[time_step, 1] <- time_step;
    sim_sum_2[time_step, 2] <- sim_new$basic_output$resource_results[1];
    sim_sum_2[time_step, 3] <- sim_new$basic_output$observation_results[1];
    sim_sum_2[time_step, 4] <- sim_new$basic_output$manager_results[3];
    sim_sum_2[time_step, 5] <- sum(sim_new$basic_output$user_results[,3]);
    sim_old                 <- sim_new;
}
colnames(sim_sum_2) <- c("Time", "Pop_size", "Pop_est", "Cull_cost", 
                         "Cull_count");
print(sim_sum_2);
```

@@

In this example, we showed how the built-in resource, observation, manager, and user sub-models can be customised by manipulating the data within the data structures that they use. The goal was to show how software users can work with these existing sub-models and data structures to customise GMSE simulations. Software users seeking even greater flexibility (e.g., replacing an entire built-in sub-model with a custom sub-model) should refer to the [Supporting Information 2](SI2.pdf) that introduces `gmse_apply` more generally. Future versions of GMSE are likely to expand on the built-in options available for simulation; requests for such expansions, or contributions, can be submitted to [GitHub](https://github.com/bradduthie/gmse/issues).

>>>>

Here we demonstrate how simulations in GMSE can be more fine-tuned to specific empirical situations through the use of `gmse_apply`. To do this, we use the same scenario described in [SI3](SI3.pdf); we first recreate the basic scenario run in `gmse` using `gmse_apply`, and then build in additional modelling details including (1) [custom placement of user land](#land), (2) [parameterisation of individual user budgets](#budgets), and (3) [density-dependent movement of resources](#movement). We emphasise that these simulations are provided only to demonstrate the use of GMSE, and specifically to show the flexibility of the `gmse_apply` function, not to accurately recreate the dynamics of a specific system or make management recommendations.

@@

Additionally, we continue to use the following values for consistency, except in the case of `stakeholders`, where we reduce the number of farmers to `stakeholders = 8`. This is done to for two reasons. First, it speeds up simulations for the purpose of demonstration; second, it makes the presentation of our custom landscape ownership easier to visualise (see below).

@@

We now recreate the simulations in [SI3](SI3.pdf), which were run using the `gmse` function, in `gmse_apply`. Doing so requires us to first initialise simulations using one call of `gmse_apply`, then loop through multiple time steps that again call `gmse_apply`; results of interest are recorded in a data frame (`sim_sum_1`). Following the protocol introduced in [SI2](SI2.pdf), we can call the initialising simulation `sim_old`, and use the code below to read in the relevant parameter values.

@@

We can then loop over 30 time steps to recreate the simulations from [SI3](SI3.pdf). In these simulations, we are specifically interested in the resource and observation outputs, as well as the manager policy and user actions for culling, which we record below in the data frame `sim_sum_1`. The inclusion of the argument `old_list` tells `gmse_apply` to use parameters and values from the list `sim_old` in the new time step.

@@

The above output from `sim_sum_1` shows the data frame that holds the information we were interested in pulling out of our simulation results. All of this information was available under the list element `sim_new$basic_output`, but other list elements of `sim_new` might also be useful to record. It is important to remember that this example of `gmse_apply` is using the default resource, observation, manager, and user sub-models. Custom sub-models could produce different outputs in `sim_new` (see [SI2](SI2.pdf) for examples). For default sub-models, there are some list elements that might be especially useful. These elements can potentially be edited *within the above loop* to dynamically adjust simulations. For more explanation of built-in GMSE data arrays, see [SI7](SI7.pdf).

- [`sim_new$resource_array`](SI7.pdf#resources): A table holding all information on resources. Rows correspond to discrete resources, and columns correspond to resource properties: (1) ID, (2-4) types (not currently in use), (5) x-location, (6) y-location, (7) movement parameter, (8) time, (9) density independent mortality parameter (`remove_pr`), (10) reproduction parameter (`lambda`), (11) offspring number, (12) age, (13-14) observation columns, (15) consumption rate (`res_consume`), and (16-20) recorded experiences of user actions (e.g., was the resource culled or scared?).
- [`sim_new$AGENTS`](SI7.pdf#agents): A table holding basic information on agents (manager and users). Rows correspond to a unique agent, and columns correspond to agent properties: (1) ID, (2) type (0 for the manager, 1 for users), (3-4) additional type options not currently in use, (5-6), x and y locations (usually ignored), (7) movement parameter (usually ignored), (8) time, (9) agent's viewing ability in cells (`agent_view`), (10) error parameter, (11-12) values for holding marks and tallies of resources, (13-15) values for holding observations, (16) yield from landscape cells, (17) budget (`manager_budget` and `user_budget`).
- `sim_new$observation_vector`: Estimate of total resource number from the observation model ([`observation_array`](SI7.pdf#observation) also holds this information in a different way depending on `observe_type`)
- [`sim_new$LAND`](SI7.pdf#land): The landscape on which interactions occur, which is stored as a 3D array with `land_dim_1` rows, `land_dim_2` columns, and 3 layers. Layer 1 (`sim_new$LAND[,,1]`) is not currently used in default sub-models, but could be used to store values that affect resources and agents. Layer 2 (`sim_new$LAND[,,2]`) stores crop yield from a cell, and layer 3 (`sim_new$LAND[,,3]`) stores the owner of the cell (value corresponds to the agent's ID).
- `sim_new$manage_vector`: The cost of each action as set by the manager. For even more fine-tuning, individual costs for the actions of each agent can be set for each user in [`sim_new$manager_array`](SI7.pdf#cost).
- `sim_new$user_vector`: The total number of actions performed by each user. A more detailed breakdown of actions by individual users is held in [`sim_new$user_array`](SI7.pdf#action).

@@

The number of cells owned by the manager (1) and each farmer (2-8) is therefore listed in the table below.

@@

```{r, echo = TRUE, eval = TRUE}
# First initialise a simulation
sim_old  <- gmse_apply(get_res = "Full", remove_pr = 0.122, lambda = 0.275,
                       res_death_K = 93870, RESOURCE_ini = 35000, 
                       manage_target = 70000, res_death_type = 3,
                       manager_budget = 10000, user_budget = 10000,
                       public_land = 0.4, stakeholders = 8, res_consume = 0.02, 
                       res_birth_K = 200000, land_ownership = TRUE,
                       observe_type = 3, agent_view = 1, converge_crit = 0.01,
                       ga_mingen = 200, res_move_type = 0);
# By setting `res_move_type = 0`, no resource movement will occur in gmse_apply
# Adjust the landscape ownership below
sim_old$LAND[1:20,   1:20,  3] <- 2; 
sim_old$LAND[1:20,  21:40,  3] <- 3;
sim_old$LAND[1:20,  41:60,  3] <- 4;
sim_old$LAND[1:20,  61:80,  3] <- 5;
sim_old$LAND[1:20,  81:100, 3] <- 6;
sim_old$LAND[21:40,  1:50,  3] <- 7;
sim_old$LAND[21:40, 51:100, 3] <- 8;
sim_old$LAND[41:60,  1:100, 3] <- 9;
sim_old$LAND[61:100, 1:100, 3] <- 1;
# Change the budgets of each farmer based on the land they own
for(ID in 2:9){
    cells_owned            <- sum(sim_old$LAND[,,3] == ID);
    sim_old$AGENTS[ID, 17] <- 100 * cells_owned;
}
# Begin simulating time steps for the system
sim_sum_2 <- matrix(data = NA, nrow = 30, ncol = 5);
for(time_step in 1:30){
    # Apply the new movement rules at the beginning of the loop
    sim_old$resource_array  <- avoid_aggregation(sim_old$resource_array);
    # Next, move on to simulate (old_list remembers that res_move_type = 0)
    sim_new                 <- gmse_apply(get_res = "Full", old_list = sim_old);
    sim_sum_2[time_step, 1] <- time_step;
    sim_sum_2[time_step, 2] <- sim_new$basic_output$resource_results[1];
    sim_sum_2[time_step, 3] <- sim_new$basic_output$observation_results[1];
    sim_sum_2[time_step, 4] <- sim_new$basic_output$manager_results[3];
    sim_sum_2[time_step, 5] <- sum(sim_new$basic_output$user_results[,3]);
    sim_old                 <- sim_new;
}
colnames(sim_sum_2) <- c("Time", "Pop_size", "Pop_est", "Cull_cost", 
                         "Cull_count");
print(sim_sum_2);
```

@@

In this example, we showed how the built-in resource, observation, manager, and user sub-models can be customised by manipulating the data within the data structures that they use. The goal was to show how software users can work with these existing sub-models and data structures to customise GMSE simulations. Readers seeking even greater flexibility (e.g., replacing an entire built-in sub-model with a custom sub-model) should refer to [SI2](SI2.pdf) that introduces `gmse_apply` more generally. Future versions of GMSE are likely to expand on the built-in options available for simulation; requests for such expansions, or contributions, can be submitted to [GitHub](https://github.com/bradduthie/gmse/issues).

### 2018-05-10 notebook/ms/SI5.Rmd
Minor edits, new read

---
title: "Fisheries example integrating FLR"
author: A. Bradley Duthie&#xb9;&#xb3;, Jeremy J. Cusack&#xb9;, Isabel L. Jones&#xb9;,
  Jeroen Minderman&#xb9;, Erlend B. Nilsen&#xb2;, Roc&#0237;o A. Pozo&#xb9;, O. Sarobidy
  Rakotonarivo&#xb9;, Bram Van Moorter&#xb2;, and Nils Bunnefeld&#xb9;
date: '[1] Biological and Environmental Sciences, University of Stirling, Stirling,
  UK [2] Norwegian Institute for Nature Research, Trondheim, Norway [3] alexander.duthie@stir.ac.uk'
output:
  pdf_document:
    citation_package: natbib
    fig_caption: yes
  html_document: default
  word_document:
    fig_caption: yes
    reference_docx: docx_template.docx
header-includes: null
linestretch: 1
link-citations: yes
linkcolor: blue
bibliography: gmse.bib
subtitle: 'GMSE: an R package for generalised management strategy evaluation (Supporting
  Information 5)'
biblio-style: apalike
---

@@

Early development of management strategy evaluation (MSE) models originated in fisheries [@Polacheck1999; @Smith1999; @Sainsbury2000]. Consequently, fisheries-focused software for MSE has been extensively developed, including R libraries that focus on the management of species of exceptional interest, such as the Atlantic Bluefin Tuna (*Thunnus thynnus*) [[ABFTMSE](https://github.com/ICCAT/abft-mse/tree/master/R_package/ABTMSE); @Carruthers2018; @Carruthers2018a], and Indian Ocean Bigeye (*T. obesus*) and Yellowfin (*T. albacares*) Tuna [[MSE-IO-BET-YFT](https://github.com/pjumppanen/MSE-IO-BET-YFT); @Kolody2016]. The largest of all such libraries is the Fisheries Library in R ([FLR](http://www.flr-project.org/)), which includes an extensive collection of tools targeted for fisheries science. The FLR library has been used in over a hundred [publications](http://www.flr-project.org/#publications) [recent publications include @Jardim2018; @Mackinson2018; @Utizi2018], and includes [an MSE framework](http://www.flr-project.org/doc/An_introduction_to_MSE_using_FLR.html) for evaluating different harvest control rules.

@@

The FLR toolset includes [a series of pacakges](http://www.flr-project.org/#packages), with several [tutorials](http://www.flr-project.org/doc/index.html) for using them. For simplicity, we focus on [a model of stock recruitment](http://www.flr-project.org/doc/Modelling_stock_recruitment_with_FLSR.html) to be used as the population model in `gmse_apply`. This population model will use sample data and one of the many available stock-recruitment models available in FLR, and a custom function will be written to return a single value for stock recruitment. Currently, `gmse_apply` requires that sub-models return subfunction results either as scalar values or data frames that are structured [in the same way](SI7.pdf) as GMSE sub-models. But interpretation of scalar values is left up to the user (e.g., population model results could be interpreted as abundance or biomass; manager policy could be interpreted as cost of harvesting or as total allowable catch). For simplicity, the observation (i.e., estimation) model will be the stock reported from the population model with error, and the manager model will be a total allowable catch calculated from the stock-recruitment relationship that accounts for the number of fishers in the system. The user model, however, will employ the full power of the default GMSE user function to simulate user actions. We first show how a custom function can be made that applies the FLR toolset to a population model.

@@

```{r, eval = FALSE}
install.packages(c("FLCore"), repos="http://flr-project.org/R");
```

@@

An initial run of these models gives values of `r round(new_rec, digits = 2)` for `new_rec` and `r round(obs_rec, digits = 2)` for `obs_rec`. We are now ready to use the built-in manager and user sub-models in `gmse_apply`. We will assume that managers attempt to keep a recruitment of 5000, and that there are 4 independent fishers [stakeholders in gmse_apply says 10, is that a difference?] who attempt to maximise their catch. We assign a user budget of `manager_budget = 10000`, and all other values are set to GMSE defaults. In the built-in GMSE functions, the manager will use the estimate of recruitment based on `obs_rec` and use it to set the cost of harvesting (`culling` in GMSE).

@@

We are now ready to loop the `gmse_apply` function over multiple time steps. To do this, we will update the `rec.m` and `ssb.m` matrices after each time step, simulating 20 years into the future. The population model `predict_recruitment` will use these data to dynamically update parameters of the Ricker model, as might occur in an empirical fishery that is being monitored. We will use the results from the observation model to update recruitment for the new year in `rec.m`. For simplicity, spawning stock biomass prior to harvest will be randomly sampled from a value in the last 10 years (i.e., from `ssb.m` between 1994 and 2004), but more realistic models could relate this spawning stock biomass to recruitment and environmental variables from a prevoius year; spawning stock biomass will be decreased after harvest based on user actions. The GMSE initialisation and simulation is below.

>>>>

---
title: "Fisheries example integrating FLR"
author: A. Bradley Duthie&#xb9;&#xb3;, Jeremy J. Cusack&#xb9;, Isabel L. Jones&#xb9;,
  Jeroen Minderman&#xb9;, Erlend B. Nilsen&#xb2;, Roc&#0237;o A. Pozo&#xb9;, O. Sarobidy
  Rakotonarivo&#xb9;, Bram Van Moorter&#xb2;, and Nils Bunnefeld&#xb9;
date: '[1] Biological and Environmental Sciences, University of Stirling, Stirling,
  UK [2] Norwegian Institute for Nature Research, Trondheim, Norway [3] alexander.duthie@stir.ac.uk'
output:
  pdf_document:
    citation_package: natbib
    fig_caption: yes
  html_document: default
  word_document:
    fig_caption: yes
    reference_docx: docx_template.docx
header-includes:
- \usepackage{lineno}
- \linenumbers
linestretch: 1
link-citations: yes
linkcolor: blue
bibliography: gmse.bib
subtitle: 'GMSE: an R package for generalised management strategy evaluation (Supporting
  Information 5)'
biblio-style: apalike
---

@@

Early development of management strategy evaluation (MSE) models originated in fisheries [@Polacheck1999; @Smith1999; @Sainsbury2000]. Consequently, fisheries-focused software for MSE has been extensively developed, including R libraries that focus on the management of species of exceptional interest, such as the Atlantic Bluefin Tuna (*Thunnus thynnus*) [[ABFTMSE](https://github.com/ICCAT/abft-mse/tree/master/R_package/ABTMSE); @Carruthers2018a; @Carruthers2018], and Indian Ocean Bigeye (*T. obesus*) and Yellowfin (*T. albacares*) Tuna [[MSE-IO-BET-YFT](https://github.com/pjumppanen/MSE-IO-BET-YFT); @Kolody2016]. The largest of all such libraries is the Fisheries Library in R ([FLR](http://www.flr-project.org/)), which includes an extensive collection of tools targeted for fisheries science. The FLR library has been used in over a hundred [publications](http://www.flr-project.org/#publications) [recent publications include @Jardim2018; @Mackinson2018; @Utizi2018], and includes [an MSE framework](http://www.flr-project.org/doc/An_introduction_to_MSE_using_FLR.html) for evaluating different harvest control rules.

@@

The FLR toolset includes [a series of packages](http://www.flr-project.org/#packages), with several [tutorials](http://www.flr-project.org/doc/index.html) for using them. For simplicity, we focus on [a model of stock recruitment](http://www.flr-project.org/doc/Modelling_stock_recruitment_with_FLSR.html) to be used as the population model in `gmse_apply`. This population model will use sample data and one of the many available stock-recruitment models available in FLR, and a custom function will be written to return a single value for stock recruitment. Currently, `gmse_apply` requires that sub-models return subfunction results either as scalar values or data frames that are structured [in the same way](SI7.pdf) as GMSE sub-models. But interpretation of scalar values is left up to the user (e.g., population model results could be interpreted as abundance or biomass; manager policy could be interpreted as cost of harvesting or as total allowable catch). For simplicity, the observation (i.e., estimation) model will be the stock reported from the population model with error. The manager and user models, however, will employ the full power of the default GMSE functions to simulate management and user actions. We first show how a custom function can be made that applies the FLR toolset to a population model.

@@

```{r, eval = FALSE}
install.packages("FLCore", repos="http://flr-project.org/R");
```

@@

An initial run of these models gives values of `r round(new_rec, digits = 2)` for `new_rec` and `r round(obs_rec, digits = 2)` for `obs_rec`. We are now ready to use the built-in manager and user sub-models in `gmse_apply`. We will assume that managers attempt to keep a recruitment of 5000, and that there are 10 independent fishers [stakeholders in gmse_apply says 10, is that a difference?] who attempt to maximise their catch. We assign a user budget of `manager_budget = 10000`, and all other values are set to GMSE defaults. In the built-in GMSE functions, the manager will use the estimate of recruitment based on `obs_rec` and use it to set the cost of harvesting (`culling` in GMSE).

@@

We are now ready to loop the `gmse_apply` function over multiple time steps. To do this, we will update the `rec.m` and `ssb.m` matrices after each time step, simulating 20 years into the future. The population model `predict_recruitment` will use these data to dynamically update parameters of the Ricker model, as might occur in an empirical fishery that is being monitored. We will use the results from the observation model to update recruitment for the new year in `rec.m`. For simplicity, spawning stock biomass prior to harvest will be randomly sampled from a value in the last 10 years (i.e., from `ssb.m` between 1994 and 2004), but more realistic models could relate this spawning stock biomass to recruitment and environmental variables from a previous year; spawning stock biomass will be decreased after harvest based on user actions. The GMSE initialisation and simulation is below.
